\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-07-09 scratch/notes_john/JSS-article/JSS-article-4.Rmd
further edits to JSS-article-4

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.

<!-- [^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions. -->

@@

[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

@@

The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner. Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model. In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression."
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```
A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths."
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "Oth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
}
m <- update(m.auto, data=A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
```{r Auto-CR-plots-transformed}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)
```

@@

The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):

@@

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
                                "MonthlyIncome")],
              y = attrition$Attrition,
              method = "glm",
              trControl = trainControl(method="LOOCV")),
  direct={
    n <- nrow(attrition)
    yhat <- numeric(n)
    for (i in 1:n){
      m <- update(mod.attrition, data=attrition[-i, ])
      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
    }
    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

>>>>

print(microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10,
  unit = "relative"
  ), signif = 3)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times=100} repetitions.

@@

[^cv-list]: In order to reduce between-model random variability in comparisons of models on the same data set, we applied `cv()` to a list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

@@

The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner. Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package, transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model.[^adding-model-selection]

[^adding-model-selection]: In a vignette on extending the \pkg{cv} package, we explain how to specify additional model-selection procedures.

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore} (a), produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the numeric variables in the \\code{Auto} data. In each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression. (b) Component+residual plots for the working model fit to the \\code{Auto} data. The broken blue lines show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.",
#| fig.subcap=rep("", 2)
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")

m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```

```{r Auto-explore-show, eval=FALSE}
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
                  smooth=list(spread=FALSE), data=Auto, pch=".")
```

A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders} (ranging from 3 to 8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model, eval=FALSE}
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-explore} (b), produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda=powers[pred])
}
m <- update(m.auto, data=A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore} (a), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (a)):
```{r Auto-transformed, fig.show="hold", echo=FALSE}
#| out.width = "45%",
#| fig.height = 5,
#| fig.cap = "(a) Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data. (b) Component+residual plots for the working model fit to the transformed \\code{Auto} data.",
#| fig.subcap=rep("", 2)
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
crPlots(m, pch=".")
```

```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore} (b), on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed} (b)):
```{r Auto-CR-plots-transformed, eval=FALSE}
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r stepAIC}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
brief(m.step)
```

@@

The other arguments to \code{cv()} are (see \code{?cv.function} for additional optional arguments and details):

@@

print(microbenchmark::microbenchmark(
  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
                                "MonthlyIncome")],
              y = attrition$Attrition,
              method = "glm",
              trControl = trainControl(method="LOOCV")),
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  direct={
    n <- nrow(attrition)
    yhat <- numeric(n)
    for (i in 1:n){
      m <- update(mod.attrition, data=attrition[-i, ])
      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
    }
    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10,
  unit = "relative"
  ), signif=3)
```
Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but almost two orders of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.
